---
title: "Coursera Crime Visualization"
author: "Javad Hashtroudian"
date: "October 10, 2017"
geometry: "left=1cm,right=1cm,top=2cm,bottom=2cm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load needed libraries

## load libraries and set working directory and clean workspace

```{r cars}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)

setwd("c:/assignment6")
rm(list = ls())
```


## load both csvs
and do a little EDA (expolrotory data analysis)


```{r}
sanf_14 <- read.csv("c:/assignment6/sanfrancisco_incidents_summer_2014.csv", stringsAsFactors = FALSE, header = TRUE )
seat_14 <- read.csv("c:/assignment6/seattle_incidents_summer_2014.csv", stringsAsFactors = FALSE, header = TRUE)

head(sanf_14)
head(seat_14)

str(sanf_14)
str(seat_14)


summary(sanf_14)
summary(seat_14)

```

## Note the two cities report their crime statistics in differen ways
Therefore I decided to examine only one city
But I will chose which city

## Next I decided to do some very preliminary plots on the cities to decide what to examine with more details


```{r}
#change SF category from string to categorical so I can plot
sanf_14$Category <- as.factor(sanf_14$Category)
plot(sanf_14$Category)
title(main = "Plot of different categories of crime in San Fransisco 2014")

```


```{r}
#change SF day of week from string to categorical so I can plot
sanf_14$DayOfWeek <- as.factor(sanf_14$DayOfWeek)
plot(sanf_14$DayOfWeek)
title(main = "Plot of crime in San Fransisco 2014 by day of week")
```

### Finally a quick look at one column from Seatle before I make my mind up

```{r}
#change from string to categorical so I can plot
seat_14$Offense.Type <- as.factor(seat_14$Offense.Type)
plot(seat_14$Offense.Type)
title(main = "Plot of different categories of crime in Seatle 2015")
```


# Based on the above preliminary data analytics I decided to examine the San Fransisco data in more detail.


```{r}
# first question is crime in Sanfransisco more prevelant at different times?

sanf_14$Time <- as.factor(sanf_14$Time)
plot(sanf_14$Time)
title(main = "Plot of crime in San Fransisco 2014 by raw time value as reported")
```


It appears that crime goes down after 10 pm to a minimum around 4am
However the above graph may well be misleading as there are two many categories

So next I will look at just the hours
```{r}
# reload the Sanfransisco data into a different data from to look at time in different ways
sf_14 <- read.csv("c:/assignment6/sanfrancisco_incidents_summer_2014.csv", stringsAsFactors = FALSE, header = TRUE )
sanf_14$timeinhour <- substring(sf_14$Time,0,2)
sanf_14$timeinhour <- as.factor(sanf_14$timeinhour)
plot(sanf_14$timeinhour)
title(main = "Plot of crime in San Fransisco 2014 by hour")

```

Definitely it appears to confirm previous idea that that crime goes down after 10 pm to a minimum around 4am
Also shows a peack at midday




Now I'm wondering if different crimes vary by time

```{r}
arson <- filter(sanf_14, sf_14$Category == 'ARSON')
plot(arson$timeinhour)
title(main = "Plot of crime of arson in San Fransisco 2014 by hour")


```


There is not enough arson data to het a good satistic for now.
So next I looked at crimes with many instances

```{r}
summary(sanf_14$Category)
```

So I'll look at LARCENY/THEFT
```{r}
lt <- filter(sanf_14, sf_14$Category == 'LARCENY/THEFT')
plot(lt$timeinhour)
title(main = "Plot of crime of larceny or theft in San Fransisco 2014 by hour")

```
```{r}
plot(lt$DayOfWeek)
title(main = "Plot of crime of larceny or theft in San Fransisco 2014 by day of week")
```

Note this case the specifc (LARCENY/THEFT) crime looks very similar to the totality of crime

So now I'm curious about same plots for VEHICLE THEFT

```{r}
vt <- filter(sanf_14, sf_14$Category == 'VEHICLE THEFT')
plot(vt$timeinhour)
title(main = "Plot of crime of vehicle theft in San Fransisco 2014 by hour")

```

```{r}
plot(vt$DayOfWeek)
title(main = "Plot of crime of vehicle theft in San Fransisco 2014 by day of week")
```

Next do the same for ASSAULT

```{r}
assult <- filter(sanf_14, sf_14$Category == 'ASSAULT')
plot(assult$timeinhour)
title(main = "Plot of crime of assult in San Fransisco 2014 by hour")

```

```{r}
plot(assult$DayOfWeek)
title(main = "Plot of crime of assult in San Fransisco 2014 by day of week")
```


It is interesting that assults continue to be high until 2am
My guess is that this relates to people drinking alcohol
There are not enough DRIVING UNDER THE INFLUENC to do any corrolation so I look up typical times bars close in San Fransisco
At https://www.tripadvisor.com/ShowTopic-g60713-i30-k1347482-Time_that_bars_close-San_Francisco_California.html
I discover that 2am is when they stop serving liquor
Also:
Most hotels' bars close at 01:00.
Most bars with entertainment close at 01:30.
Most gin mills close at 01:50.
Last call for alcohol is usually twenty minutes before closing.

So my visualization made me predict a corralation between time of assults and bar closing times


